|

### 2025-01-31 doc/paper.qmd
more

---
title: |
  | How Much New Information Is There in Earnings?
  | Corporate Decision-Making and Quantitative Analysis
  | Winter 2024/25 - Individual Report\vspace{1cm}
author:
  - name: Melisa Mazaeva
    email: melisa.mazaeva@student.hu-berlin.de
    affiliations:
      - Humboldt-Universität zu Berlin  
date: today
date-format: MMM D, YYYY [\vspace{0cm}] # Adjusted the vertical space after the date from 1cm
abstract: |
  | This project uses the TRR 266 Template for Reproducible Empirical Accounting Research (TREAT) to provide infrastructure for open science-oriented empirical projects. Leveraging data from the CRSP and Compustat databases via WRDS, alongside Worldscope and Datastream via Thomson/Refinitiv, this repository showcases a reproducible workflow integrating Python scripts for data preparation, analysis, and visualization. Integrating multiple databases adds complexity, requiring a detailed understanding of their structures and careful scripting to extract, align, and analyze data effectively. The project replicates and extends Ball and Shivakumar (2008) to analyze the informativeness of quarterly earnings announcements and their contribution to annual share price movements, highlighting their critical role in investment decisions and impact on investors, analysts, and policymakers. Key tasks include replicating and comparing original results, extending the analysis to 2007–2023, and applying the methodology to a non-U.S. country. The project also documents research design choices, discusses variations between original and reproduced results, and provides insights into earnings informativeness across different timeframes and jurisdictions. Additionally, it sketches a research design for a non-archival study to evaluate the paper’s findings. This code base, adapted from TREAT, demonstrates how the template applies to this project and serves as a structured guide for reproducible empirical research in accounting.
  <!--
  | \vspace{6cm} commenting away to decrease distance
  -->

@@

# Introduction {#sec-introduction}
The aim of this paper is to illustrate the use of open science tools in empirical accounting research. This project builds on the methodological framework established in the Corporate Decision-Making and Quantitative Analysis (CQA) course, which explored a range of empirical research methods—including archival analysis, field experiments, and survey-based approaches—to develop a comprehensive understanding of corporate decision-making and quantitative analysis. Expanding on prior empirical research conducted in Assignments I and II, this paper extends the empirical focus of the course. Assignment I examined audit market concentration in the EU using Transparency Reports data, offering insights into the dominance of major audit firms and the structure of the European audit market. Assignment II explored the economic implications of corporate financial reporting, focusing on @graham_economic_2005 to analyze managerial decision-making, earnings management strategies, and voluntary disclosure practices. This study highlighted the trade-offs between short-term earnings predictability and long-term firm value and provided a foundation for understanding corporate incentives in financial reporting.

Building on these foundations, Assignment III investigates the informativeness of quarterly earnings announcements and their contribution to annual share price movements. The project replicates and extends @ball_how_2008, assessing the extent to which earnings releases serve as a source of new information for financial markets. By integrating multiple datasets from WRDS and applying an event-study methodology, this study evaluates the economic role of earnings disclosures in market efficiency, showcasing the valuation purpose of accounting. At the same time, earnings reports may play a crucial role in contracting by influencing managerial compensation, debt agreements, and other contractual mechanisms tied to financial performance [@ball_how_2008]. The integration of multiple databases adds complexity, requiring a detailed understanding of their structures and careful scripting to extract, align, and analyze data effectively. Key tasks include replicating and comparing original results, extending the analysis to data from 2007–2023, and applying the methodology to Canada as a non-U.S. country to explore cross-market variations.

Beyond empirical replication, the project also documents research design choices, replication steps, and explicit assumptions made whenever the original paper was unclear on how to proceed. It discusses variations between original and reproduced results and provides insights into the timeliness and market impact of earnings announcements across different timeframes and jurisdictions. Additionally, it sketches a research design for a non-archival study to evaluate the paper’s findings through an alternative methodological lens. This project leverages the TRR 266 Template for Reproducible Empirical Accounting Research (TREAT) to establish an open science-oriented infrastructure, ensuring transparency, replicability, and structured workflows for empirical accounting research.

Earnings announcements provide investors with valuable information about a firm’s market value, with stock prices reacting significantly when earnings news deviates from expectations [@fink_review_2021, p.2]. The seminal study by @ball_empirical_1968 was the first to document this relationship, showing that stock prices anticipate earnings surprises, with most of the market reaction occurring before the official announcement, suggesting that earnings reports primarily confirm rather than introduce new information. Over the past decades, more than a thousand studies have examined the interplay between capital markets and financial statements, a research stream that originated with Ball and Brown (1968)  [@kothari_capital_2001]. This reinforces the role of earnings announcements as a crucial source of financial information.

Building on this foundation, the original study by @ball_how_2008 investigates the extent to which quarterly earnings announcements contribute new information to the market, assessing their role in shaping annual share price movements. By estimating the R² from regressions of annual stock returns on earnings announcement window returns, the paper quantifies the informativeness of earnings releases, finding that they account for only 1% to 2% of total annual volatility. This challenges the assumption that earnings provide substantial new information and instead suggests a confirmatory role in financial reporting, reinforcing prior market expectations rather than acting as a primary source of new insights. Their findings align with the semi-strong form of market efficiency, which suggests that stock prices already reflect all publicly available information, including prior signals about earnings. Unlike traditional event studies that assume market efficiency, @ball_how_2008 allow for potential market mispricing by not imposing a unit slope restriction in their regressions, meaning that their approach can capture under- or overreaction to earnings announcements. They also emphasize that extending the event window beyond the earnings announcement period would shift the research focus from earnings “surprise” to broader financial reporting effects, such as its role in debt contracting and executive compensation. The study further documents an increasing trend in earnings informativeness in recent years, potentially linked to regulatory changes, shifts in analyst activity, or broader market conditions.

While @ball_how_2008 provide key insights into the informativeness of earnings announcements, their study was conducted within a specific timeframe and market context. Since then, financial markets have undergone significant changes due to regulatory reforms, economic crises, and technological advancements in financial information processing. Moreover, the extent to which earnings informativeness varies across international markets remains underexplored. By replicating their study with updated data from 2007–2023 and applying the methodology to a non-U.S. country, this project reassesses the robustness of their findings, examines long-term trends, and evaluates cross-country differences in earnings informativeness. Through this approach, it critically evaluates the generalizability of the original results and provides further insights into the economic role of earnings announcements in financial markets, offering a contemporary perspective on their evolving informativeness and implications for investors, analysts, and policymakers.

The paper is structured into sections corresponding to Tasks 1–3 (@sec-task1, @sec-task2, @sec-task3), each detailing the research design choices and assumptions, documenting the replication steps, and analyzing the replication results for the respective segment. @sec-task1 follows an exact replication approach by replicating key tables and figures from the original study using US data (1972–2006) to establish baseline results. @sec-task2 extends the analysis through empirical generalization, applying the same methodology to 2007–2023 US data to compare how results evolve over time. @sec-task3 employs a generalization and extension approach by applying the methodology to a non-US country, requiring adjustments for different databases and market structures. This framework aligns with the replication taxonomy outlined by @salterio_replication_2022. Finally, @sec-task4 sketches the survey design as a non-archival study that allows evaluating the key findings of the seminal paper, following Bloomfield’s (2016) triangulation principle. The concluding remarks are provided in @sec-conclusion.  The entire Python computation code for regression run is available in `code/python/do_analysis.py` and `code/python/do_analysis-wscp.py` for detailed review if necessary. This paper focuses on the replication process, presenting visualizations, and discussing the results.

>>>>

---
title: |
  | How Much New Information Is There in Earnings?
  | Corporate Decision-Making and Quantitative Analysis
  | Winter 2024/25 - Individual Report\vspace{1cm}
author:
  - name: Melisa Mazaeva
    email: melisa.mazaeva@student.hu-berlin.de
    affiliations:
      - Humboldt-Universität zu Berlin  
date: today
date-format: MMM D, YYYY [\vspace{0cm}] # Adjusted the vertical space after the date from 1cm
abstract: |
  | This project uses the TRR 266 Template for Reproducible Empirical Accounting Research (TREAT) to provide infrastructure for open science-oriented empirical projects. Leveraging data from the CRSP and Compustat databases, alongside Worldscope and Datastream via WRDS, this repository showcases a reproducible workflow integrating Python scripts for data preparation, analysis, and visualization. Integrating multiple databases adds complexity, requiring a detailed understanding of their structures and careful scripting to extract, align, and analyze data effectively. The project replicates and extends Ball and Shivakumar (2008) to analyze the informativeness of quarterly earnings announcements and their contribution to share price movements, highlighting their critical role in investment decisions and impact on investors and analysts. Key tasks include replicating and comparing original results, extending the analysis to 2007–2023, and applying the methodology to a non-U.S. country. The project also documents research design choices, discusses variations between original and reproduced results, and provides insights into earnings informativeness across different timeframes and jurisdictions. Additionally, it sketches a research design for a non-archival study to evaluate the paper’s findings. This code base, adapted from TREAT, demonstrates how the template applies to this project and serves as a structured guide for reproducible empirical research in accounting.
  <!--
  | \vspace{6cm} commenting away to decrease distance
  -->

@@

# Introduction {#sec-introduction}
The aim of this paper is to illustrate the use of open science tools in empirical accounting research. This project builds on the methodological framework established in the Corporate Decision-Making and Quantitative Analysis (CQA) course, which explored a range of empirical research methods—including archival analysis, field experiments, and survey-based approaches—to develop a comprehensive understanding of corporate decision-making and quantitative analysis. Expanding on prior empirical research conducted in Assignments I and II, this paper extends the empirical focus of the course. Assignment I examined audit market concentration in the EU using Transparency Reports data, offering insights into the dominance of major audit firms and the structure of the European audit market. Assignment II explored the economic implications of corporate financial reporting, focusing on @graham_economic_2005 to analyze managerial decision-making, earnings management strategies, and voluntary disclosure practices.

Building on these foundations, Assignment III investigates the informativeness of quarterly earnings announcements and their contribution to annual share price movements. The project replicates and extends @ball_how_2008, assessing the extent to which earnings releases serve as a source of new information for financial markets. By integrating multiple datasets from WRDS and applying an event-study methodology, this study aims to evaluate the role of earnings disclosures in market efficiency, showcasing the valuation purpose of accounting. At the same time, earnings reports may play a crucial role in contracting by influencing managerial compensation, debt agreements, and other contractual mechanisms tied to financial performance [@ball_how_2008]. The integration of multiple databases adds complexity, requiring a detailed understanding of their structures and careful scripting to extract, align, and analyze data effectively.

Beyond empirical replication, the project also documents research design choices, replication steps, and explicit assumptions made whenever the original paper was unclear on how to proceed. It discusses variations between original and reproduced results and provides insights into the timeliness and market impact of earnings announcements across different timeframes and jurisdictions. Additionally, it sketches a research design for a non-archival study to evaluate the paper’s findings through an alternative methodological lens. This project leverages the TRR 266 Template for Reproducible Empirical Accounting Research (TREAT) to establish an open science-oriented infrastructure, ensuring transparency and replicability for empirical accounting research.

Earnings announcements provide investors with valuable information about a firm’s market value, with stock prices reacting significantly when earnings news deviate from expectations [@fink_review_2021, p.2]. The seminal study by @ball_empirical_1968 was the first to document this relationship, showing that stock prices anticipate earnings surprises, with most of the market reaction occurring before the official announcement, suggesting that earnings reports primarily confirm rather than introduce new information. Over the past decades, more than a thousand studies have examined the interplay between capital markets and financial statements, a research stream that originated with Ball and Brown (1968)  [@kothari_capital_2001].

Building on this foundation, the original study by @ball_how_2008 investigates the extent to which quarterly earnings announcements contribute new information to the market, assessing their role in shaping annual share price movements. By estimating the R² from regressions of annual stock returns on earnings announcement window returns, the paper quantifies the informativeness of earnings releases, finding that they account for only 1% to 2% of total annual volatility. This challenges the assumption that earnings provide substantial new information and instead suggests a confirmatory role in financial reporting. These findings align with the semi-strong form of market efficiency, which suggests that stock prices already reflect all publicly available information, including prior signals about earnings. Unlike traditional event studies that assume market efficiency, @ball_how_2008 allow for potential market mispricing by not imposing a unit slope restriction in their regressions, meaning that their approach can capture under- or overreaction to earnings announcements. They also emphasize that extending the event window beyond the earnings announcement period would shift the research focus from earnings “surprise” to broader financial reporting effects, such as its role in debt contracting and executive compensation.

While @ball_how_2008 provide key insights into the informativeness of earnings announcements, their study was conducted within a specific timeframe and market context. Since then, financial markets have undergone significant changes due to regulatory reforms, economic crises, and technological advancements in financial information processing. Moreover, the extent to which earnings informativeness varies across international markets remains underexplored. By replicating their study with updated data from 2007–2023 and applying the methodology to Canada as a non-U.S. country, this project reassesses the robustness of their findings. Through this approach, it critically evaluates the generalizability of the original results.

The paper is structured into sections corresponding to Tasks 1–3 (@sec-task1, @sec-task2, @sec-task3), each detailing the research design choices and assumptions, documenting the replication steps and results. @sec-task1 follows an exact replication approach by replicating key tables and figures from the original study using US data (1972–2006) to establish baseline results. @sec-task2 extends the analysis through empirical generalization, applying the same methodology to 2007–2023 US data to compare how results evolve over time. @sec-task3 employs a generalization and extension approach by applying the methodology to a non-US country, requiring adjustments for different databases and market structures. This framework aligns with the replication taxonomy outlined by @salterio_replication_2022. Finally, @sec-task4 sketches the survey design as a non-archival study that allows evaluating the key findings of the seminal paper, following Bloomfield’s (2016) triangulation principle. The concluding remarks are provided in @sec-conclusion.
